Responsible Machine Learning Techniques\\\large{Interpretable Models, Post-hoc Explanation, and Disparate Impact Testing}

### 2019-12-04 article-information-2019.tex
progress on figures, tables, discussion, and intro

ML models can be inaccurate and unappealable black-boxes, even with the application of newer post-hoc explanation techniques \cite{please_stop}.\footnote{See: \href{https://www.nytimes.com/2017/06/13/opinion/how-computers-are-harming-criminal-justice.html}{``When a Computer Program Keeps You in Jail''}.} ML models can perpetuate and exacerbate discrimination \cite{fairmlbook}, \cite{feldman2015certifying}, \cite{dwork2012fairness}, \cite{gender_shades}.  ML models can be hacked, resulting in manipulated model outcomes or the exposure of proprietary intellectual property or sensitive training data \cite{security_of_ml}, \cite{model_stealing}, \cite{membership_inference}. This text makes no claim that the interdependent issues of opaqueness, discrimination, or security vulnerabilities in ML have been solved, even as singular entities, much less as complex intersectional phenomena, e.g. the fairwashing or scaffolding of biased models with ML explanations or the privacy harms of ML explanations \cite{fair_washing}, \cite{scaffolding}, \cite{shokri2019privacy}. This text does however propose some workable countermeasures to address these vexing problems in the form of interpretable models, post-hoc explanation, and DI and discrimination testing implemented in widely available, free, and open source Python tools. Section \ref{sec:m_and_m} describes methods and materials, including simulated and collected training datasets, interpretable and constrained model architectures, post-hoc explanations used to create an \textit{appealable} decision-making framework, tests for DI and other discrimination, and public and open source software resources. In Section \ref{sec:res}, interpretable and constrained modeling results are compared to less interpretable and unconstrained models and post-hoc explanation and discrimination testing results are also presented for interpretable models. Section \ref{sec:disc} then discusses some nuances of the outlined modeling, explanation, and discrimination testing methods and results. Finally, Section \ref{sec:con} closes this text with a brief outline of proposed additional steps to increase human trust and understanding in ML.

@@

@@

Because unconstrained ML models, like gradient boosting machines (GBMs) (e.g. \cite{friedman2001greedy}, \cite{esl}) and artificial neural networks (ANNs) (e.g. \cite{recht2011hogwild}, \cite{hinton2012improving}, \cite{sutskever2013importance}, \cite{zeiler2012adadelta}), can be difficult to understand, trust, and appeal, even after the application of post-hoc explanation techniques, explanation analysis and discrimination testing are applied only to the constrained interpretable ML models \cite{please_stop}, \cite{fair_washing}, \cite{scaffolding}. Here, MGBMs\footnote{As implemented in \href{https://xgboost.readthedocs.io/en/latest/tutorials/monotonic.html}{\texttt{XGBoost}} or \href{https://github.com/h2oai/h2o-3/blob/master/h2o-py/demos/H2O_tutorial_gbm_monotonicity.ipynb}{\texttt{h2o}}.} and XNNs (\cite{wf_xnn} \cite{yang2019enhancing}) will serve as those more interpretable models for subsequent explanatory and discrimination analysis. Presented explanation techniques include PD, ICE, and Shapley values \cite{esl}, \cite{ice_plots}, \cite{shapley}, \cite{tree_shap}. PD, ICE, and Shapley values provide direct, global, and local summaries and descriptions of constrained models without resorting to the use of intermediary and approximate surrogate models. \textcolor{red}{Discussed discrimination testing methods and metrics include ... }. All outlined materials and methods are implemented in open source Python code, and are made available in the software resources associated with this text. Detailed descriptions of notation, training data, ML models, post-hoc explanation techniques, discrimination testing methods, and software resources are organized in Section \ref{sec:m_and_m} as follows:

\begin{itemize}
	\item \textbf{Notation}: spaces, datasets, \& models -- \textsection\ref{ssec:not}
	\item \textbf{Training data}: simulated data \& collected mortgage data -- \textsection\ref{ssec:sim_data}~and~\textsection\ref{ssec:mort_data}
	\item \textbf{ML models}: constrained, interpretable MGBM \& XNN models  -- \textsection\ref{ssec:mgbm}~and~\textsection\ref{ssec:xnn}
	\item \textbf{Post-hoc explanation techniques}: PD, ICE, \& Shapley values -- \textsection\ref{ssec:pd_ice}~and~\textsection\ref{ssec:shap}
	\item \textbf{Discrimination testing methods}: -- \textsection\ref{ssec:di}
	\item \textbf{Software resources}: GitHub repository for this text; utilized \& useful Python packages -- \textsection\ref{ssec:soft}
\end{itemize}

@@

	\begin{itemize}
		\item Input features come from the set $\mathcal{X}$ contained in a \textit{P}-dimensional input space, $\mathcal{X} \subset \mathbb{R}^P$.  An arbitrary, potentially unobserved, or future instance of $\mathcal{X}$ is denoted $\mathbf{x}$, $\mathbf{x} \in \mathcal{X}$.
		\item Labels corresponding to instances of $\mathcal{X}$ come from the set $\mathcal{Y}$.
		\item Learned output responses of models are contained in the set $\mathcal{\hat{Y}}$. % For regression models, the set $\mathcal{\hat{Y}}_r$ is also contained in a $C$-dimensional output space, $\mathcal{\hat{Y}}_r \subset \mathbb{R}^{C_r}$. For classification models, the set $\mathcal{\hat{Y}}_c$ typically contains a column vector for each unique class in $\mathcal{Y}$. Hence, $\mathcal{\hat{Y}}_c$ is contained in a $C'$-dimensional output space,  $\mathcal{\hat{Y}}_c \subset \mathbb{R}^{C'_c}$.
	\end{itemize}

@@

	\begin{itemize}
		\item The input dataset $\mathbf{X}$ is composed of observed instances of the set $\mathcal{X}$ with a corresponding dataset of labels $\mathbf{Y}$, observed instances of the set $\mathcal{Y}$. 
		\item Each $i$-th observation of $\mathbf{X}$ is denoted as $\mathbf{x}^{(i)} = $  
		$[x_0^{(i)}, x_1^{(i)}, \dots, x_{\textit{P}-1}^{(i)}]$, with corresponding $i$-th labels in $\mathbf{Y}, \mathbf{y}^{(i)}$, and corresponding predictions in $\mathbf{\hat{Y}}, \mathbf{\hat{y}}^{(i)}$. % = [y_0^{(i)}, y_1^{(i)}, \dots, y_{\textit{C}-1}^{(i)}]$.
		\item $\mathbf{X}$ and $\mathbf{Y}$ consist of $N$ tuples of observations: $[(\mathbf{x}^{(0)},\mathbf{y}^{(0)}), (\mathbf{x}^{(1)},\mathbf{y}^{(1)}), \dots,(\mathbf{x}^{(N-1)},\mathbf{y}^{(N-1)})]$. %\\ $\mathbf{x}^{(i)} \in \mathcal{X}$, $\mathbf{y}^{(i)} \in \mathcal{Y}$.
		\item Each $j$-th input column vector of $\mathbf{X}$ is denoted as $X_j = [x_{j}^{(0)}, x_{j}^{(1)}, \dots, x_{j}^{(N-1)}]^T$.
	\end{itemize}

@@

	\begin{itemize}
		\item A type of ML model $g$, selected from a hypothesis set $\mathcal{H}$, is trained to represent an unknown signal-generating function $f$ observed as  $\mathbf{X}$ with labels $\mathbf{Y}$ using a training algorithm $\mathcal{A}$: 
		$ \mathbf{X}, \mathbf{Y} \xrightarrow{\mathcal{A}} g$, such that $g \approx f$.
		\item $g$ generates learned output responses on the input dataset $g(\mathbf{X}) = \mathbf{\hat{Y}}$, and on the general input space $g(\mathcal{X}) = \mathcal{\hat{Y}}$.
		\item The model to be explained and tested for discrimination testing is denoted as $g$.
	\end{itemize}

@@

\begin{enumerate}
\item For the first and highest split in $T_b$ involving $X_j$, any $\theta_{b,j,0}$ resulting in the left child weight being greater than the right child weight, $T(x_j; \theta_{b,j,0}) = \{w_{b,j,0,L}, w_{j,0,R}\}$ where $w_{b,j,0,L} > w_{b,j,0,R}$, is not considered. 
\item For any subsequent left child node involving $X_j$, any $\theta_{b,j, k\ge1}$ resulting in $T(x_j; \theta_{b,j,k\ge1}) = \{w_{b,j,k\ge1,L}, w_{b,j,k\ge1,R}\}$ where $w_{b,j,k\ge1,L} > w_{b,j,k\ge1,R}$, is not considered.
\item Moreover, for any subsequent left child node involving $X_j$, $T(x_j; \theta_{b,j,k\ge1}) = \{w_{b,j,k\ge1,L}, w_{b,j,k\ge1,R}\}$, $\{w_{b,j,k\ge1,L}, w_{b,j,k\ge1,R}\}$ are bound by the parent set of node weights, $\{w_{b,j,k-1,L}, w_{b,j,k-1, R}\}$, such that $ \{w_{b,j,k\ge1,L}, w_{b,j,k\ge1,R}\} \le \frac{w_{b,j,k-1,L} + w_{b,j,k-1,R}}{2}$.
\item (1) and (2) are also applied to all right child nodes, except that for right child nodes $\{w_{b,j,k\ge1,L}, w_{b,j,k\ge1,R}\} \ge \frac{w_{b,j,k-1,L} + w_{b,j,k-1,R}}{2}$.
\end{enumerate}

@@

\begin{enumerate}
\item The first and deepest meta-layer, composed of $K$ linear $\sum_j\beta_{k,j}x_j$ hidden units, is known as the \textit{projection layer} and is fully connected to each input feature, $X_j$.
\item The second meta-layer contains $K$ hidden and separate $n_k$ ridge functions, or \textit{subnetworks}. Each $n_k$ is a neural network, which can be parameterized to suite a given modeling task. To facilitate direct visualization, the input to each subnetwork is the 1-dimensional output of its associated projection layer hidden unit, $\sum_j\beta_{k,j}x_j$.
\item The output meta-layer, called the \textit{combination layer}, is another linear unit comprised of a global bias term, $\mu_0$, and the $K$ weighted 1-dimensional outputs of each subnetwork, $\gamma_kn_k(\sum_j\beta_{k,j}x_j)$. Again, subnetwork output is restricted to 1-dimension for visualization purposes.
\end{enumerate}

@@

Shapley values can be estimated in different ways, many of which are intractable for datasets with large $\mathcal{P}$. Tree SHAP is a specific implementation of Shapley explanations that relies on traversing internal decision tree structures to efficiently estimate the contribution of each $x_j$ for some $g(\mathbf{x})$ \cite{tree_shap}. Tree SHAP has been implemented efficiently in popular gradient boosting libraries such as \href{http://docs.h2o.ai/h2o/latest-stable/h2o-py/docs/modeling.html#h2ogradientboostingestimator}{\texttt{h2o}}, \href{https://lightgbm.readthedocs.io/en/latest/Python-Intro.html}{\texttt{LightGBM}}, and \href{https://xgboost.readthedocs.io/en/latest/python/python_intro.html}{\texttt{XGBoost}}, and Tree SHAP is used to calculate accurate and consistent global and local feature importance for MGBM models in this text. Unfortunately, many non-consistent explanation methods can result in drastically different global and local feature importance values across different models trained on the same data or even for refreshing the same model with augmented training data \cite{molnar}. Consistency and accuracy guarantees are perhaps a factor in the growing momentum behind Shapley values as a candidate technique for generating consumer-specific and personalized adverse action notices for automated ML-based decisions in highly-regulated settings such as credit lending \cite{bracke2019machine}.

@@

Python code to reproduce discussed results is available at: \texttt{\url{https://github.com/h2oai/article-information-2019}}. The primary Python packages employed therein are:

\begin{itemize}
	\item \href{https://numpy.org/}{\texttt{numpy}} and \href{https://pandas.pydata.org/}{\texttt{pandas}} for data manipulation.
	\item \href{https://github.com/h2oai/h2o-3}{\texttt{h2o}}, \href{https://keras.io/}{\texttt{keras}}, \href{https://github.com/slundberg/shap}{\texttt{shap}}, and \href{https://www.tensorflow.org/}{\texttt{tensorflow}} for modeling, explanation, and discrimination testing.
	\item \href{https://matplotlib.org/}{\texttt{matplotlib}} and \href{https://seaborn.pydata.org/}{\texttt{seaborn}} for plotting. 
\end{itemize}

Additional, relevant Python packages include: \href{https://github.com/dssg/aequitas}{\texttt{aequitas}} and \href{https://github.com/LASER-UMASS/Themis}{\texttt{Themis}} for discrimination testing, \href{http://aif360.mybluemix.net/}{\texttt{AIF360}} for discrimination testing and remediation, \href{https://github.com/interpretml/interpret}{\texttt{interpret}} for interpretable models and post-hoc explanation, and \href{https://github.com/SeldonIO/alibi}{\texttt{alibi}} and \href{https://github.com/SauceCat/PDPbox}{\texttt{PDPbox}} for post-hoc explanation.\footnote{See: \texttt{\url{https://github.com/jphall663/awesome-machine-learning-interpretability}} for a longer, curated list of related software packages and resources.}

@@

\subsubsection{Unconstrained Model Fit Assessment}

\subsubsection{Constrained Model Fit Assessment}

@@

@@

\subsubsection{Unconstrained Model Fit Assessment}

\subsubsection{Constrained Model Fit Assessment}

@@

@@

\subsection{The Burgeoning Ecosystem of Interpretable Machine Learning Models}

MGBM and XNN architectures were selected for this text because they are straightforward variants of popular unconstrained ML models. If practitioners are working with GBM and ANN models, it should be relatively uncomplicated for them to evaluate the constrained versions of these models. The same can be said for the discussed post-hoc explanation and discrimination testing techniques. Practitioners should be able to augment many interpretable modeling workflows with Shapley and derivative-based post-hoc explanatory techniques and \textcolor{red}{...} discrimination testing methods. These approaches are promising responses to the opaqueness and discrimination problems in ML. However, they are imperfect and just a small part of a larger and burgeoning responsible ML ecosystem.

% GA2M, Rudin Group

@@

%appeal

@@

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\funding{This research received no external funding.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
\abbreviations{The following abbreviations are used in this text: ANN -- artificial neural network, DI -- disparate impact, GBM -- gradient boosting machine, ICE -- individual conditional expectation, MGBM -- monotonic gradient boosting machine, ML -- machine learning, PD -- partial dependence, SGD -- stochastic gradient descent, US -- United States, XNN -- explainable neural network.}

>>>>

Responsible artificial intelligence (AI) has been variously conceptualized as AI-based products or projects that use transparent technical mechanisms, that create appealable decisions or outcomes, that perform reliably and in a trustworthy manner over time, that exhibit minimal social discrimination, and that are designed by humans with diverse experiences, both in terms of demographics and professional backgrounds, i.e. ethics, social sciences, and technology.\footnote{See: \href{https://ec.europa.eu/jrc/communities/sites/jrccties/files/03_dignum_v.pdf}{Responsible Artificial Intelligence}, \href{https://www.accenture.com/_acnmedia/pdf-92/accenture-afs-responsible-ai.pdf}{\textit{Responsible AI: A Framework for Building Trust in Your AI Solutions}}, \href{https://www.pwc.com/us/en/services/consulting/analytics/artificial-intelligence/what-is-responsible-ai.html}{PwC's Responsible AI}, \href{https://ai.google/responsibilities/responsible-ai-practices/}{Responsible AI Practices}} Even if responsible AI feels like a somewhat broad and amorphous notion, at least one aspect should be crystal clear: ML models, a common application of AI, have problems that responsible practitioners should likely attempt to remediate. ML models can be inaccurate and unappealable black-boxes, even with the application of newer post-hoc explanation techniques \cite{please_stop}.\footnote{See: \href{https://www.nytimes.com/2017/06/13/opinion/how-computers-are-harming-criminal-justice.html}{``When a Computer Program Keeps You in Jail''}.} ML models can perpetuate and exacerbate discrimination \cite{fairmlbook}, \cite{feldman2015certifying}, \cite{dwork2012fairness}, \cite{gender_shades}.  ML models can be hacked, resulting in manipulated model outcomes or the exposure of proprietary intellectual property or sensitive training data \cite{security_of_ml}, \cite{model_stealing}, \cite{membership_inference}. Though this text makes no claim that the interdependent issues of opaqueness, discrimination, or security vulnerabilities in ML have been solved (even as singular entities, much less as complex intersectional phenomena), this text does propose some specific technical countermeasures, in the form of interpretable models, post-hoc explanation, and DI and discrimination testing implemented in widely available, free, and open source Python tools, to address a subset of these vexing problems.

Section \ref{sec:m_and_m} describes methods and materials, including simulated and collected training datasets, interpretable and constrained model architectures, post-hoc explanations used to create an \textit{appealable} decision-making framework, tests for DI and other discrimination, and public and open source software resources. In Section \ref{sec:res}, interpretable and constrained modeling results are compared to less interpretable and unconstrained models and post-hoc explanation and discrimination testing results are also presented for interpretable models. Section \ref{sec:disc} then discusses some nuances of the outlined modeling, explanation, and discrimination testing methods and results. Section \ref{sec:con} closes this text with a brief outline of proposed additional steps to increase human trust and understanding in ML.

@@

Detailed descriptions of notation, training data, ML models, post-hoc explanation techniques, discrimination testing methods, and software resources are organized in Section \ref{sec:m_and_m} as follows:

\begin{itemize}[leftmargin=*,labelsep=5.8mm]
	\item \textbf{Notation}: spaces, datasets, \& models -- \textsection\ref{ssec:not}
	\item \textbf{Training data}: simulated data \& collected mortgage data -- \textsection\ref{ssec:sim_data}~and~\textsection\ref{ssec:mort_data}
	\item \textbf{ML models}: constrained, interpretable MGBM \& XNN models  -- \textsection\ref{ssec:mgbm}~and~\textsection\ref{ssec:xnn}
	\item \textbf{Post-hoc explanation techniques}: PD, ICE, \& Shapley values -- \textsection\ref{ssec:pd_ice}~and~\textsection\ref{ssec:shap}
	\item \textbf{Discrimination testing methods}: -- \textsection\ref{ssec:di}
	\item \textbf{Software resources}: GitHub repository for this text; utilized \& useful Python packages -- \textsection\ref{ssec:soft}
\end{itemize}

@@

Because unconstrained ML models, like gradient boosting machines (GBMs) (e.g. \cite{friedman2001greedy}, \cite{esl}) and artificial neural networks (ANNs) (e.g. \cite{recht2011hogwild}, \cite{hinton2012improving}, \cite{sutskever2013importance}, \cite{zeiler2012adadelta}), can be difficult to understand, trust, and appeal, even after the application of post-hoc explanation techniques, explanation analysis and discrimination testing are applied only to the constrained interpretable ML models \cite{please_stop}, \cite{fair_washing}, \cite{scaffolding}. Here, MGBMs\footnote{As implemented in \href{https://xgboost.readthedocs.io/en/latest/tutorials/monotonic.html}{\texttt{XGBoost}} or \href{https://github.com/h2oai/h2o-3/blob/master/h2o-py/demos/H2O_tutorial_gbm_monotonicity.ipynb}{\texttt{h2o}}.} and XNNs (\cite{wf_xnn} \cite{yang2019enhancing}) will serve as those more interpretable models for subsequent explanatory and discrimination analysis. Presented explanation techniques include PD, ICE, and Shapley values \cite{esl}, \cite{ice_plots}, \cite{shapley}, \cite{tree_shap}. PD, ICE, and Shapley values provide direct, global, and local summaries and descriptions of constrained models without resorting to the use of intermediary and approximate surrogate models. \textcolor{red}{Discussed discrimination testing methods and metrics include ... }. All outlined materials and methods are implemented in open source Python code, and are made available in the software resources associated with this text.

@@

	\begin{itemize}[leftmargin=*,labelsep=5.8mm]
		\item Input features come from the set $\mathcal{X}$ contained in a \textit{P}-dimensional input space, $\mathcal{X} \subset \mathbb{R}^P$.  An arbitrary, potentially unobserved, or future instance of $\mathcal{X}$ is denoted $\mathbf{x}$, $\mathbf{x} \in \mathcal{X}$.
		\item Labels corresponding to instances of $\mathcal{X}$ come from the set $\mathcal{Y}$.
		\item Learned output responses of models are contained in the set $\mathcal{\hat{Y}}$. % For regression models, the set $\mathcal{\hat{Y}}_r$ is also contained in a $C$-dimensional output space, $\mathcal{\hat{Y}}_r \subset \mathbb{R}^{C_r}$. For classification models, the set $\mathcal{\hat{Y}}_c$ typically contains a column vector for each unique class in $\mathcal{Y}$. Hence, $\mathcal{\hat{Y}}_c$ is contained in a $C'$-dimensional output space,  $\mathcal{\hat{Y}}_c \subset \mathbb{R}^{C'_c}$.
	\end{itemize}

@@

	\begin{itemize}[leftmargin=*,labelsep=5.8mm]
		\item The input dataset $\mathbf{X}$ is composed of observed instances of the set $\mathcal{X}$ with a corresponding dataset of labels $\mathbf{Y}$, observed instances of the set $\mathcal{Y}$. 
		\item Each $i$-th observation of $\mathbf{X}$ is denoted as $\mathbf{x}^{(i)} = $  
		$[x_0^{(i)}, x_1^{(i)}, \dots, x_{\textit{P}-1}^{(i)}]$, with corresponding $i$-th labels in $\mathbf{Y}, \mathbf{y}^{(i)}$, and corresponding predictions in $\mathbf{\hat{Y}}, \mathbf{\hat{y}}^{(i)}$. % = [y_0^{(i)}, y_1^{(i)}, \dots, y_{\textit{C}-1}^{(i)}]$.
		\item $\mathbf{X}$ and $\mathbf{Y}$ consist of $N$ tuples of observations: $[(\mathbf{x}^{(0)},\mathbf{y}^{(0)}), (\mathbf{x}^{(1)},\mathbf{y}^{(1)}), \dots,(\mathbf{x}^{(N-1)},\mathbf{y}^{(N-1)})]$. %\\ $\mathbf{x}^{(i)} \in \mathcal{X}$, $\mathbf{y}^{(i)} \in \mathcal{Y}$.
		\item Each $j$-th input column vector of $\mathbf{X}$ is denoted as $X_j = [x_{j}^{(0)}, x_{j}^{(1)}, \dots, x_{j}^{(N-1)}]^T$.
	\end{itemize}

@@

	\begin{itemize}[leftmargin=*,labelsep=5.8mm]
		\item A type of ML model $g$, selected from a hypothesis set $\mathcal{H}$, is trained to represent an unknown signal-generating function $f$ observed as  $\mathbf{X}$ with labels $\mathbf{Y}$ using a training algorithm $\mathcal{A}$: 
		$ \mathbf{X}, \mathbf{Y} \xrightarrow{\mathcal{A}} g$, such that $g \approx f$.
		\item $g$ generates learned output responses on the input dataset $g(\mathbf{X}) = \mathbf{\hat{Y}}$, and on the general input space $g(\mathcal{X}) = \mathcal{\hat{Y}}$.
		\item The model to be explained and tested for discrimination testing is denoted as $g$.
	\end{itemize}

@@

\begin{enumerate}[leftmargin=*,labelsep=4.9mm]
\item For the first and highest split in $T_b$ involving $X_j$, any $\theta_{b,j,0}$ resulting in the left child weight being greater than the right child weight, $T(x_j; \theta_{b,j,0}) = \{w_{b,j,0,L}, w_{j,0,R}\}$ where $w_{b,j,0,L} > w_{b,j,0,R}$, is not considered. 
\item For any subsequent left child node involving $X_j$, any $\theta_{b,j, k\ge1}$ resulting in $T(x_j; \theta_{b,j,k\ge1}) = \{w_{b,j,k\ge1,L}, w_{b,j,k\ge1,R}\}$ where $w_{b,j,k\ge1,L} > w_{b,j,k\ge1,R}$, is not considered.
\item Moreover, for any subsequent left child node involving $X_j$, $T(x_j; \theta_{b,j,k\ge1}) = \{w_{b,j,k\ge1,L}, w_{b,j,k\ge1,R}\}$, $\{w_{b,j,k\ge1,L}, w_{b,j,k\ge1,R}\}$ are bound by the parent set of node weights, $\{w_{b,j,k-1,L}, w_{b,j,k-1, R}\}$, such that $ \{w_{b,j,k\ge1,L}, w_{b,j,k\ge1,R}\} \le \frac{w_{b,j,k-1,L} + w_{b,j,k-1,R}}{2}$.
\item (1) and (2) are also applied to all right child nodes, except that for right child nodes $\{w_{b,j,k\ge1,L}, w_{b,j,k\ge1,R}\} \ge \frac{w_{b,j,k-1,L} + w_{b,j,k-1,R}}{2}$.
\end{enumerate}

@@

\begin{enumerate}[leftmargin=*,labelsep=4.9mm]
\item The first and deepest meta-layer, composed of $K$ linear $\sum_j\beta_{k,j}x_j$ hidden units, is known as the \textit{projection layer} and is fully connected to each input feature, $X_j$.
\item The second meta-layer contains $K$ hidden and separate $n_k$ ridge functions, or \textit{subnetworks}. Each $n_k$ is a neural network, which can be parameterized to suite a given modeling task. To facilitate direct visualization, the input to each subnetwork is the 1-dimensional output of its associated projection layer hidden unit, $\sum_j\beta_{k,j}x_j$.
\item The output meta-layer, called the \textit{combination layer}, is another linear unit comprised of a global bias term, $\mu_0$, and the $K$ weighted 1-dimensional outputs of each subnetwork, $\gamma_kn_k(\sum_j\beta_{k,j}x_j)$. Again, subnetwork output is restricted to 1-dimension for visualization purposes.
\end{enumerate}

@@

Shapley values can be estimated in different ways, many of which are intractable for datasets with large $\mathcal{P}$. Tree SHAP is a specific implementation of Shapley explanations that relies on traversing internal decision tree structures to efficiently estimate the contribution of each $x_j$ for some $g(\mathbf{x})$ \cite{tree_shap}. Tree SHAP (SHapley Additive exPlanations) has been implemented efficiently in popular gradient boosting libraries such as \href{http://docs.h2o.ai/h2o/latest-stable/h2o-py/docs/modeling.html#h2ogradientboostingestimator}{\texttt{h2o}}, \href{https://lightgbm.readthedocs.io/en/latest/Python-Intro.html}{\texttt{LightGBM}}, and \href{https://xgboost.readthedocs.io/en/latest/python/python_intro.html}{\texttt{XGBoost}}, and Tree SHAP is used to calculate accurate and consistent global and local feature importance for MGBM models in this text. Unfortunately, many non-consistent explanation methods can result in drastically different global and local feature importance values across different models trained on the same data or even for refreshing the same model with augmented training data \cite{molnar}. Consistency and accuracy guarantees are perhaps a factor in the growing momentum behind Shapley values as a candidate technique for generating consumer-specific and personalized adverse action notices for automated ML-based decisions in highly-regulated settings such as credit lending \cite{bracke2019machine}.

@@

Python code to reproduce discussed results is available at: \texttt{\url{https://github.com/h2oai/article-information-2019}}. The primary Python packages employed are: \href{https://numpy.org/}{\texttt{numpy}} and \href{https://pandas.pydata.org/}{\texttt{pandas}} for data manipulation, \href{https://github.com/h2oai/h2o-3}{\texttt{h2o}}, \href{https://keras.io/}{\texttt{keras}}, \href{https://github.com/slundberg/shap}{\texttt{shap}}, and \href{https://www.tensorflow.org/}{\texttt{tensorflow}} for modeling, explanation, and discrimination testing, and \href{https://matplotlib.org/}{\texttt{matplotlib}} and \href{https://seaborn.pydata.org/}{\texttt{seaborn}} for plotting.

@@

\subsubsection{Constrained vs. Unconstrained Model Fit Assessment}

\begin{table}[H]
\label{tab:sim_acc}
\caption{Accuracy metrics for $g^\text{GBM}$, $g^\text{MGBM}$, $g^\text{ANN}$, and $g^\text{XNN}$ on simulated data.}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{ccccc}
\toprule
\textbf{Model}		& \textbf{Accuracy}	& \textbf{AUC}	 & \textbf{Logloss}	& \textbf{RMSE}\\
\midrule
$g^\text{GBM}$		& 				& 			& 				& \\
$g^\text{MGBM}$	& 				& 			& 				& \\
$g^\text{ANN}$		& 				& 			& 				& \\
$g^\text{XNN}$		& 				& 			& 				& \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{figure}[H]
\centering
%\includegraphics[width=8cm]{img/}
\label{fig:sim_mgbm_glob}
\caption{PD and ICE across deciles for the five most important input features for $g^\text{MGBM}(\mathbf{X})$ on the test partition of the simulated data.}
\end{figure}

\begin{figure}[H]
\centering
%\includegraphics[width=8cm]{img/}
\label{fig:sim_mgbm_loc}
\caption{Mean Tree SHAP values across quintiles for the five most important input features for $g^\text{MGBM}(\mathbf{X})$ on the test partition of the simulated data.}
\end{figure}

\begin{figure}[H]
\centering
%\includegraphics[width=8cm]{img/}
\label{fig:sim_xnn_glob}
\caption{Ridge functions \textcolor{red}{and ICE across deciles} for the five most important input features for $g^\text{XNN}(\mathbf{X})$ on the test partition of the simulated data.}
\end{figure}

\begin{figure}[H]
\centering
%\includegraphics[width=8cm]{img/}
\label{fig:fig:sim_xnn_glob}
\caption{Mean \textcolor{red}{local feature importance} values across quintiles for the five most important input features for $g^\text{XNN}(\mathbf{X})$ on the test partition of the simulated data.}
\end{figure}

@@

\subsubsection{Constrained vs. Unconstrained Model Fit Assessment}

\begin{table}[H]
\label{tab:mort_acc}
\caption{Accuracy metrics for $g^\text{GBM}$, $g^\text{MGBM}$, $g^\text{ANN}$, and $g^\text{XNN}$ on mortgage data.}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{ccccc}
\toprule
\textbf{Model}		& \textbf{Accuracy}	& \textbf{AUC}	 & \textbf{Logloss}	& \textbf{RMSE}\\
\midrule
$g^\text{GBM}$		& 				& 			& 				& \\
$g^\text{MGBM}$	& 				& 			& 				& \\
$g^\text{ANN}$		& 				& 			& 				& \\
$g^\text{XNN}$		& 				& 			& 				& \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{figure}[H]
\centering
%\includegraphics[width=8cm]{img/}
\label{fig:mort_mgbm_glob}
\caption{PD and ICE across deciles for the five most important input features for $g^\text{MGBM}(\mathbf{X})$ on the test partition of the mortgage data.}
\end{figure}

\begin{figure}[H]
\centering
%\includegraphics[width=8cm]{img/}
\label{fig:mort_mgbm_loc}
\caption{Mean Tree SHAP values across quintiles for the five most important input features for $g^\text{MGBM}(\mathbf{X})$ on the test partition of the mortgage data.}
\end{figure}

\begin{figure}[H]
\centering
%\includegraphics[width=8cm]{img/}
\label{fig:mort_xnn_glob}
\caption{Ridge functions \textcolor{red}{and ICE across deciles} for the five most important input features for $g^\text{XNN}(\mathbf{X})$ on the test partition of the mortgage data.}
\end{figure}

\begin{figure}[H]
\centering
%\includegraphics[width=8cm]{img/}
\label{fig:fig:mort_xnn_glob}
\caption{Mean \textcolor{red}{local feature importance} values across quintiles for the five most important input features for $g^\text{XNN}(\mathbf{X})$ on the test partition of the mortgage data.}
\end{figure}

@@

\subsection{The Burgeoning Python Ecosystem for Responsible Machine Learning}

MGBM and XNN interpretable model architectures were selected for this text because they are straightforward variants of popular unconstrained ML models. If practitioners are working with GBM and ANN models, it should be relatively uncomplicated for them to evaluate the constrained versions of these models. These approaches are promising responses to the black-box problem in ML. However, they are just a small part of a burgeoning ecosystem of Python tools for interpretable models, post-hoc explanation, and discrimination testing and remediation. Any discussion of interpretable ML models would be incomplete without references to the seminal work of the Rudin group at Duke University and EBM or GA\textsuperscript{2}M models, pioneered by researchers Microsoft and Cornell and led by Rich Caruana. In keeping with a major theme of this text, models from these leading researchers and several other kinds of interpretable ML models are now available as open source Python packages. Among others, practitioners can now evaluate EBM in the \href{https://github.com/interpretml/interpret}{\texttt{interpret}} package, optimal sparse decision trees, GAMs in the \href{https://github.com/dswah/pyGAM}{\texttt{pyGAM}} package, a variant of Friedman's RuleFit  in the \href{https://github.com/scikit-learn-contrib/skope-rules}{\texttt{skope-rules}} package, monotonic calibrated interpolated lookup tables in \href{https://github.com/tensorflow/lattice}{\texttt{tensorflow/lattice}}, and ``this looks like that'' interpretable deep learning \cite{osdt}, \cite{rulefit}, \cite{lattice}, \cite{this_looks_like_that}.\footnote{Optimal sparse decision trees: \url{https://github.com/xiyanghu/OSDT}.}\textsuperscript{,}\footnote{``This looks like that'' interpretable deep learing: \url{https://github.com/cfchen-duke/ProtoPNet}.} Additional, relevant Python packages include: \href{https://github.com/dssg/aequitas}{\texttt{aequitas}} and \href{https://github.com/LASER-UMASS/Themis}{\texttt{Themis}} for discrimination testing, \href{http://aif360.mybluemix.net/}{\texttt{AIF360}} for discrimination testing and remediation, and \href{https://github.com/SeldonIO/alibi}{\texttt{alibi}} and \href{https://github.com/SauceCat/PDPbox}{\texttt{PDPbox}} for post-hoc explanation.\footnote{See: \texttt{\url{https://github.com/jphall663/awesome-machine-learning-interpretability}} for a longer, curated list of related software packages and resources.}

@@

%appeal: interpretable models + explanations

@@

% e.g. the fairwashing or scaffolding of biased models with ML explanations or the privacy harms of ML explanations \cite{fair_washing}, \cite{scaffolding}, \cite{shokri2019privacy})

@@

\begin{figure}[H]
\centering
\includegraphics[width=8cm]{img/blueprint.png}
\label{fig:blueprint}
\caption{A diagram of a proposed holistic ML workflow in which explanations (highlighted in red) are used along with interpretable models, DI analysis and remediation techniques, and other review and appeal mechanisms to create an understandable and trustworthy ML system.}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\funding{This work received no external funding.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
\abbreviations{The following abbreviations are used in this text: AI -- artificial intelligence, ANN -- artificial neural network, DI -- disparate impact, EBM or GA\textsuperscript{2}M -- explainable boosting machine, i.e. variants GAMs that consider two-way interactions and incorporate boosting into training, GAM -- generalized additive model, GBM -- gradient boosting machine, ICE -- individual conditional expectation, MGBM -- monotonic gradient boosting machine, ML -- machine learning, PD -- partial dependence, SGD -- stochastic gradient descent, US -- United States, XNN -- explainable neural network.}
